---
title: "Exploration of the Mtcars dataset"
author: "Hélène Descamps"
date: "2025-06-14"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load libraries
```{r, message=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(smplot2)
library(ggpubr)
```

## load and print statistics of mtcars
```{r, message=FALSE}
data(mtcars)
```
```{r}
summary(mtcars)
#sapply(mtcars, sd)
dim(mtcars) 
colnames(mtcars) 
head(mtcars, 5)
```

## Principal component analysis
```{r, message=FALSE}
pca.res <- prcomp(mtcars, scale.=F, retx=T)
pc.var<-pca.res$sdev^2 
pc.per<-round(pc.var/sum(pc.var)*100, 1) 
pca.res.df <- as_tibble(pca.res$x)%>%
  mutate(car=rownames(pca.res$x))

rotation<-as_tibble(pca.res$rotation)%>%
  mutate(features=rownames(pca.res$rotation))%>%
  pivot_longer(!features,names_to="PC")

```

```{r}
ggplot(data= filter(rotation, PC %in% c("PC1", "PC2", "PC3", "PC4")) , aes(x=features, y=value))+
  geom_point()+
  facet_wrap(~PC, scales="free")+
  theme(strip.background = element_blank(), axis.text.x=element_text(angle=90))+
  theme(text=element_text(size=20))
#ggsave("PCvsFeatures.tiff")
```
```{r}
pca.res.df<-merge(pca.res.df, mtcars, by.x="car", by.y="row.names")
ggplot(pca.res.df, aes(x=PC1, y=PC2))+
  geom_point(aes(color=disp, size=hp)) +
  ylim(-200,200)+
  xlab(paste0("PC1 (",pc.per[1],"%",")"))+
  ylab(paste0("PC2 (",pc.per[2],"%",")"))+
  labs(color="disp", size="hp")+
  theme_classic()+
  theme(aspect.ratio = 1.25)+
  theme(text=element_text(size=20))
#ggsave("PC1dispVsPC2hp.tiff")
```

> Displacement is the primary contributor to PC1, which explains 92.7% of the variance in the mtcars dataset, while horsepower predominantly influences PC2, accounting for an additional 7.2% of the total variance.

## How does weight influences miles/gallon?
```{r, message=FALSE}
ggplot(mtcars, aes(x=mpg, y=wt))+
  geom_point()+
  sm_statCorr(corr_method='pearson', R2=TRUE, color="darkblue" ,label_x=20, label_y=5)+
  theme_classic()+
  labs(x="Miles per gallon", y="Weight (1000lbs)")+
  theme(aspect.ratio = 1.25)+
  theme(text=element_text(size=20))
#ggsave("MPGvsWeight.tiff")
```

> Miles per gallon is inversely correlated to the vehicle weight, which alone accounts for 75% of the variation in fuel efficiency.

## What is the average mpg by cyl ?
```{r}
mtcars%>%
  group_by(cyl)%>%
  summarise(mean_mpg=mean(mpg))%>%
  ggplot(aes(x=as.factor(cyl), y=mean_mpg, fill=cyl))+
  geom_col(show.legend=FALSE)+
  labs(x="cylinder", y="miles per gallon (average)")+
  theme_classic()+
  theme(aspect.ratio = 1.25)+
  theme(text=element_text(size=20))
#ggsave("meanMPGvsCyl.tiff")
```

## Create a linear model of mpg by wt and hp
```{r}
model <- lm(mpg ~ wt+ hp, data=mtcars)
summary(model)
```

# Categorize Horsepower into high and low groups based on the median value, and perform a t-test to assess the significance of their effect on miles per gallon.
```{r}
mtcars2<-mtcars%>%
  dplyr::mutate(hp_cat=ifelse(hp<median(mtcars$hp),"low", "high"))
t.test(mtcars2$mpg ~ mtcars2$hp_cat)

compare_means(mpg ~ hp_cat, data=mtcars2, method="t.test")

ggplot(mtcars2, aes(x=factor(hp_cat, levels=c("low", "high")), y=mpg, fill=hp_cat))+
  geom_boxplot(show.legend=FALSE)+
  labs(x="horsepower", y="miles per gallon")+
  theme_classic()+
  theme(aspect.ratio = 1.25)+
  stat_compare_means(method="t.test")
#ggsave("HPhighVSlow_mpg.tiff")
```

> Higher horsepower significantly reduces miles/gallon (95% CI [-12,-6], p=1.6e-06)