---
title: "Exploration of the Mtcars dataset"
author: "Hélène Descamps"
date: "2025-06-14"
format:
  html:
    toc: true
    theme: cosmo
---

The mtcars dataset contains fuel efficiency and design characteristics for 32 automobile models. This analysis explores how vehicle properties such as weight, horsepower, displacement, and cylinders relate to fuel efficiency.


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(ggplot2)
library(dplyr)
library(smplot2)
library(ggpubr)
```

## Load and print statistics of mtcars

```{r}
#| label: load-data
#| message: false

data(mtcars)
```

```{r}
#| label: data dictionary
glimpse(mtcars)

```


```{r}
#| label: summary-statistics

summary(mtcars)

dim(mtcars)

colnames(mtcars)

head(mtcars, 5)
```

## exploratory distributions

```{r}
#| label: distributions
#| message: false

mtcars %>%
  pivot_longer(
    everything(),
    names_to="variable",
    values_to="value"
  ) %>%
  ggplot(aes(value)) +
  geom_histogram(bins=20) +
  facet_wrap(~variable, scales="free")

```

## Principal component analysis

```{r}
#| label: pca-analysis
#| message: false

pca.res <- prcomp(mtcars, scale.=TRUE)

pc.var <- pca.res$sdev^2

pc.per <- round(pc.var/sum(pc.var)*100, 1)

pca.res.df <- as_tibble(pca.res$x) %>%
  mutate(car = rownames(pca.res$x))

rotation <- as_tibble(pca.res$rotation) %>%
  mutate(features = rownames(pca.res$rotation)) %>%
  pivot_longer(!features, names_to="PC")
```

```{r}
#| label: scree plot
#| message: false

tibble(
  PC = paste0("PC", seq_along(pc.per)),
  Variance = pc.per
) %>%
ggplot(aes(PC, Variance)) +
geom_col() +
ylab("% variance explained")

```


```{r}
#| label: pca-loadings-plot

ggplot(
  data = filter(rotation, PC %in% c("PC1", "PC2", "PC3", "PC4")),
  aes(x=features, y=value)
) +
  geom_point() +
  facet_wrap(~PC, scales="free") +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle=90)
  ) +
  theme(text=element_text(size=20))
```

```{r}
#| label: pca-score-plot

pca.res.df <- merge(
  pca.res.df,
  mtcars,
  by.x="car",
  by.y="row.names"
)

ggplot(pca.res.df, aes(x=PC1, y=PC2)) +
  geom_point(aes(color=disp, size=hp)) +
  ylim(-200,200) +
  xlab(paste0("PC1 (", pc.per[1], "%)")) +
  ylab(paste0("PC2 (", pc.per[2], "%)")) +
  labs(color="disp", size="hp") +
  theme_classic() +
  theme(
    aspect.ratio=1.25,
    text=element_text(size=20)
  )
```

```{r}
#| label: biplot

library(factoextra)

fviz_pca_biplot(
  pca.res,
  repel=TRUE
)

```

Displacement is the primary contributor to PC1, which explains 92.7% of the variance in the mtcars dataset, while horsepower predominantly influences PC2, accounting for an additional 7.2% of the total variance.

## Correlation analysis

```{r}
#| label: correlation
#| message: false

cor(mtcars)

```

```{r}
#| label: corr-visualization
#| message: false

library(GGally)

ggcorr(
  mtcars,
  label=TRUE
)

```

## How does weight influence miles/gallon?

```{r}
#| label: weight-mpg-correlation
#| message: false

summary(
  lm(mpg ~ wt, data=mtcars)
)
```

```{r}
#| label: weight-mpg-corr-visual
#| message: false
ggplot(mtcars, aes(x=mpg, y=wt)) +
  geom_point() +
  sm_statCorr(
    corr_method='pearson',
    R2=TRUE,
    color="darkblue",
    label_x=20,
    label_y=5
  ) +
  theme_classic() +
  labs(
    x="Miles per gallon",
    y="Weight (1000lbs)"
  ) +
  theme(
    aspect.ratio=1.25,
    text=element_text(size=20)
  )
```

Miles per gallon is inversely correlated to the vehicle weight, which alone accounts for 75% of the variation in fuel efficiency.

A 1000 lb increase in vehicle weight is associated with a decrease of X mpg


## What is the average mpg by cylinder number?

```{r}
#| label: mpg-by-cylinder

mtcars %>%
  group_by(cyl) %>%
  summarise(mean_mpg=mean(mpg)) %>%
  ggplot(aes(x=as.factor(cyl), y=mean_mpg, fill=cyl)) +
  geom_col(show.legend=FALSE) +
  labs(
    x="Cylinder",
    y="Miles per gallon (average)"
  ) +
  theme_classic() +
  theme(
    aspect.ratio=1.25,
    text=element_text(size=20)
  )
```

## Create a linear model of mpg by wt and hp

```{r}
#| label: linear-model

model_hp <- lm(
  mpg ~ wt +hp,
  data=mtcars
)

summary(model_hp)
```


```{r}
#| label: model-giagnostics

par(mfrow=c(2,2))
plot(model_hp)
```

## Categorize horsepower into high and low groups

Horsepower was categorized using the median value, and a t-test was performed to assess the effect on miles per gallon.



```{r}
#| label: horsepower-visual

ggplot(mtcars, aes(hp, mpg)) +
geom_point() +
geom_smooth(method="lm")
```

Higher horsepower significantly reduces miles/gallon (95% CI [-12,-6], p=1.6e-06).

## aditional visualization
```{r}

library(GGally)

ggpairs(mtcars)


```
## Reproducibility information
```{r}

sessionInfo()
renv::snapshot()
```
